On Optimality of Finite Element Integration

### 2015-12-07 TOMS_Paper/assembly.tex
MAH

As demonstrated by the substantial body of research on the topic, automating the generation of such high performance implementations poses several challenges. This is a result of the complexity inherent to the mathematical expressions involved in the numerical integration, which varies from problem to problem, and the particular structure of the loop nests enclosing the integrals. General-purpose compilers, such as \emph{GNU's} and \emph{Intel's}, fail at exploiting the structure inherent in the expressions, thus producing sub-optimal code (i.e., code which performs more floating-point operations, or ``flops'', than necessary; we show this is Section~\ref{sec:perf-results}). Research compilers, for instance those based on polyhedral analysis of loop nests such as PLUTO (\cite{PLUTO}), focus on parallelization and optimization for cache locality, treating issues orthogonal to the question of minimising flops. The lack of suitable third-party tools has led to the development of a number of domain-specific code transformation (or synthesizer) systems. In~\citeN{quadrature1}, it is shown how automated code generation can be leveraged to introduce optimizations that a user should not be expected to write ``by hand''. In~\citeN{FFC-TC} and~\citeN{Francis}, mathematical reformulations of finite element integration are studied with the aim of minimizing the operation count. In~\citeN{Luporini}, the effects and the interplay of generalized code motion and a set of low level optimizations are analysed. It is also worth mentioning an on-going effort to produce a new form compiler, called UFLACS (\cite{Uflacs}), which adds to the already abundant set of code transformation systems for assembly operators. The performance evaluation in Section~\ref{sec:perf-results} includes most of these optimization systems.

@@

\begin{itemize}
\item We formalize optimality for finite element integration and we build the space of legal transformations within which the optimal is sought.
\item We provide an algorithm to select points in the transformation space. The algorithm uses a cost model to: (i) understand whether a transformation reduces or increases the operation count; (ii) choose between different (non-composable) transformations.
\item We explain under what conditions our algorithm leads to optimality. In particular, we show that (i) for a particular class of problems, optimality is reached; (ii) quasi-optimality is, at least, always achieved (i.e., the operation count is at least optimal in innermost loops).
\item We integrate our approach with a compiler, COFFEE\footnote{COFFEE stands for COmpiler For Fast Expression Evaluation. The compiler is open-source and available at \url{https://github.com/coneoproject/COFFEE}}, which is in use in the Firedrake framework.
\item We experimentally evaluate using a broader suite of forms, discretizations, and code generation systems than has been used in prior research. This is essential to demonstrate that our optimality model holds in practice.
\end{itemize}

@@

In this section, we study \textit{sharing elimination}, a transformation that aims to reduce the operation count by removing sharing through the application of factorization, generalized code motion and (if necessary) common sub-expression elimination. If the objective were reaching optimality and expressions were systematically unstructured, a transformation of this sort would require solving a large combinatorial problem -- for instance to evaluate the impact of all possible factorization strategies. Our sharing elimination strategy, instead, attempts to exploit the structure inherent in finite element integration expressions to guarantee quasi-optimality\footnote{This require coordination with other transformations, as discussed in the following sections. For the moment -- and for ease of treatise -- we neglect this aspect.}, with optimality being achieved whenever stronger conditions hold. Relaxing the problem is essential to produce simple and computationally efficient algorithms -- which, as such, can be integrated with a compiler.

In Sections~\ref{sec:se-mln} and~\ref{sec:se-rln} we discuss the structural (i.e., inherent in finite element integration) and algebraic (i.e., related to the use of differential and vector operators) properties of our expressions, while in Section~\ref{sec:sec:se-algo} we present the sharing elimination algorithm. We recall that we are assuming a generic finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$ with expressions in normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$.

\subsubsection{Common sub-expressions}
\label{sec:se-mln}
A typical optimization to reduce the arithmetic complexity of a code is common sub-expression elimination (CSE). This often revolves around searching, possibly across different statements, and replacing with temporaries, sub-expressions that (i) are symbolically identical and (ii) reduce to the same value.  We observe that ``classic CSE'' (CCSE) can straightforwardly be applied without the need for data dependence analysis, since the body of a finite element integration loop nest is an independent basic block. More interestingly, we observe that in the ``special case'' (SCSE) in which test and trial functions are in the same function space, there may exist $w_1$ and $w_2$ such that $\alpha_{eij}^{w1} = \beta_{eik}^{w2}$, in which case common sub-expressions will arise along different loops ($L_j$ and $L_k$). General-purpose compilers cannot detect this (very domain-specific) property. Our algorithm for sharing elimination will exploit this extra information. Although both CSE and factorization target spatial sharing, the latter may actually increase the operation count if preliminary transformations (e.g., expansion) are required. In such a case, CCSE will be preferred.

\subsubsection{Structured tensor operations}
\label{sec:se-rln}
Finite element expressions can be seen as composition of operations between tensors. We observe that, when multiplying two tensors, it often happens that the optimal scheduling strategy is to be searched in a space of size 2. To illustrate the concept, we take a matrix-vector product $y_{\Psi \Delta} = A_{\Psi} x_{\Delta}$, in which the entries of $y, A$ and $x$ depend on $\Psi, \Delta \subset \Lambda$. Let us assume that $y_{\Psi \Delta}$ itself is an operand of a larger scalar-valued expression, so its entries can represent any possible sub-expression in $\lbrace \alpha_{eij}, \beta_{eik}, \sigma_{ei} \rbrace$. For simplicity, assume that $A$ is the inverse Jacobian matrix of a given coefficient and $x$ the gradient of a function in two dimensions, and consider the case $\Psi = \emptyset$, $\Delta = \lbrace L_i \rbrace$. Thus an entry in $y_{L_i}$ is of the form $(a f_i + b g_i) (c f_i + d g_i) \in \sigma_{ei}$. To schedule this expression, we basically have two options:
\begin{enumerate}
\item We leave the expression unmodified in the body of $L_i$; we may want to assign it to a temporary in the case it is a common sub-expression
\item We notice the presence of spatial sharing. We can re-factorize the expression as $f_i (a + c) + g_i (b + d)$, which promotes temporal sharing and, therefore, code motion. 
\end{enumerate}
The reader can verify that option 2) reduces the operation count with respect to 1) if and only if the size of $L_i$ is strictly greater than $c=1$. One can find this dichotomy in a wide range of variational forms\footnote{In fact, it systematically appears in all of the forms used for experimentation (Section~\ref{sec:perf-results}), and many more.}, obviously with values of $c$ varying with the operational cost inherent in removing spatial sharing.

In general, given a product of two sub-expressions in a finite element integration loop nest, we observe that whenever there is no ambiguity in the elimination of spatial sharing, the problem of finding an optimal scheduling reduces to the comparison of two alternatives. We note that this condition simply holds if a loop is guaranteed to be linear; in the next section, we will show how take advantage of this property to ensure quasi-optimal sharing elimination. Further, if the $\sigma_{ei}$ expressions are sufficiently regular, then the same approach can be applied to all expressions logically belonging to $L_i$. The regularity of $\sigma_{ei}$ is not as uncommon as it may seem at first glance: for instance, this property holds for many of the tensors that need be evaluated in the complex hyperelastic model analyzed in Section~\ref{sec:perf-results} (e.g., the Cauchy-Green tensor).

@@

\subsubsection{Algorithm}
\label{sec:sec:se-algo}
Algorithm~\ref{algo:opt} details the steps for performing sharing elimination on a tree representation of the loop nest. The algorithm is meant to be applied to each statement appearing in the independent basic block of the input finite element integration loop nest.

\begin{Algo}[Sharing elimination]
\label{algo:opt}
\normalfont
The first step consists of a depth-first visit of the expression tree with recursive collection of identical sub-expressions. This is to ensure that the operands of a generic operation are unique. For example, the expression $(a + a + b + c + a)$, in which we have a single n-ary operator \texttt{+}, is transformed into $(3a + b + c)$.

We then perform a second depth-first visit of the expression tree and we recursively label each node $n$ with a couple ${<}I_{jk},\ cost{>}$, where: $I_{jk}$ represents the intersection of the symbols depending on the multilinear loop nest (i.e., depending on test and trial functions) within $n$, whereas $cost$ denotes the operation count if generalized code motion were applied to all sub-expressions rooted in $n$ (i.e., if they were placed sufficiently far in the loop nest such that redundant computation were avoided). A crucial observation is that $I_{jk}$ is, in practice, always very small (a few units), even if the form is complex and includes several monomials. In addition, its computation can be made extremely efficient by also tracking the union of the symbols rooted in $n$.

On a final depth-first visit, we examine the label of each operator \texttt{+}. If $I_{jk} = \emptyset$ -- that is, there is no spatial sharing -- we just proceed to the parent node. Otherwise, we speculatively assume that structured tensor operations are performed in the sub-tree rooted in \texttt{+}, and compare the operational cost of the two alternatives described in Section~\ref{sec:se-rln} (see points (1) and (2)). Alternative (1) has an operation count equal to $cost$. Alternative (2) requires collecting the shared symbols and estimating the impact of generalized code motion on the new factorization. For this, we need decide the collection order; that is, whether to start with symbols in $I_j$ or $I_k$. It can easily be demonstrated by contradiction that starting with $I_o$ such that $|L_o| |I_o|$ is minimum ($o \in {\lbrace} j, k{\rbrace}, |.|$ represents a generic size operator) leads to the lowest operation count. If found to be the most favorable, alternative (2) is straightforwardly applied; alternative (1) is instead applied only after all operands rooted in \texttt{+} have been examined.

>>>>

As demonstrated by the substantial body of research on the topic, automating the generation of such high performance implementations poses several challenges. This is a result of the complexity inherent in the mathematical expressions involved in the numerical integration, which varies from problem to problem, and the particular structure of the loop nests enclosing the integrals. General-purpose compilers, such as \emph{GNU's} and \emph{Intel's}, fail at exploiting the structure inherent in the expressions, thus producing sub-optimal code (i.e., code which performs more floating-point operations, or ``flops'', than necessary; we show this is Section~\ref{sec:perf-results}). Research compilers, for instance those based on polyhedral analysis of loop nests such as PLUTO (\cite{PLUTO}), focus on parallelization and optimization for cache locality, treating issues orthogonal to the question of minimising flops. The lack of suitable third-party tools has led to the development of a number of domain-specific code transformation (or synthesizer) systems. In~\citeN{quadrature1}, it is shown how automated code generation can be leveraged to introduce optimizations that a user should not be expected to write ``by hand''. In~\citeN{FFC-TC} and~\citeN{Francis}, mathematical reformulations of finite element integration are studied with the aim of minimizing the operation count. In~\citeN{Luporini}, the effects and the interplay of generalized code motion and a set of low level optimizations are analysed. It is also worth mentioning an on-going effort to produce a new form compiler, called UFLACS (\cite{Uflacs}), which adds to the already abundant set of code transformation systems for assembly operators. The performance evaluation in Section~\ref{sec:perf-results} includes most of these optimization systems.

@@

\begin{itemize}
\item We formalize finite element integration loop nests and we build the space of legal transformations impacting their operation count.
\item We provide an algorithm to select points in the transformation space. The algorithm uses a cost model to: (i) understand whether a transformation reduces or increases the operation count; (ii) choose between different (non-composable) transformations.
\item We explain under what conditions our algorithm leads to optimality. In particular, we show that (i) for a particular class of problems, optimality is reached; (ii) quasi-optimality is, in general, always achieved (i.e., the operation count is at least optimal in innermost loops).
\item We integrate our approach with a compiler, COFFEE\footnote{COFFEE stands for COmpiler For Fast Expression Evaluation. The compiler is open-source and available at \url{https://github.com/coneoproject/COFFEE}}, which is in use in the Firedrake framework.
\item We experimentally evaluate using a broader suite of forms, discretizations, and code generation systems than has been used in prior research. This is essential to demonstrate that our optimality model holds in practice.
\end{itemize}

@@

In this section, we study \textit{sharing elimination}, a transformation that aims to reduce the operation count by removing sharing through the application of factorization, generalized code motion and (if necessary) common sub-expression elimination. If the objective were reaching optimality and expressions were systematically unstructured, a transformation of this sort would require solving a large combinatorial problem -- for instance to evaluate the impact of all possible factorizations. Our sharing elimination strategy, instead, attempts to exploit the structure inherent in finite element integration expressions to guarantee quasi-optimality\footnote{This require coordination with other transformations, as discussed in the following sections. For the moment -- and for ease of treatise -- we neglect this aspect.}, with optimality being achieved if stronger hypotheses hold. Relaxing the problem is essential to produce simple and computationally efficient algorithms -- two necessary conditions for integration with a compiler.

Sections~\ref{sec:se-mln} and~\ref{sec:se-rln} discuss the structural (i.e., inherent in finite element integration) and algebraic (i.e., related to the use of differential and vector operators) properties that can characterize our expressions. Section~\ref{sec:sec:se-algo} presents the sharing elimination algorithm. We recall that we are assuming a generic finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$ with expressions in normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$.

\subsubsection{Common sub-expressions}
\label{sec:se-mln}
A typical optimization to reduce the arithmetic complexity of a code is common sub-expression elimination (CSE). This often revolves around searching, possibly across different statements, and replacing with temporaries, sub-expressions that are (i) symbolically identical and (ii) reduce to the same value.  We observe that ``classic CSE'' (CCSE) can straightforwardly be applied without the need for data dependence analysis, since the body of a finite element integration loop nest is an independent basic block. More interestingly, we observe that in the ``special case'' (SCSE) in which test and trial functions are in the same function space, there may exist $w_1$ and $w_2$ such that $\alpha_{eij}^{w1} = \beta_{eik}^{w2}$, in which case common sub-expressions will arise along different loops ($L_j$ and $L_k$). General-purpose compilers cannot detect this (very domain-specific) property. Our algorithm for sharing elimination will exploit this extra information. Although both CSE and factorization target spatial sharing, the latter may actually increase the operation count if preliminary transformations (e.g., expansion) are required. In such a case, CCSE will be preferred.

\subsubsection{Structured tensor operations}
\label{sec:se-rln}
Finite element expressions can be seen as composition of operations between tensors. We observe that, when multiplying two tensors, it often happens that the optimal scheduling strategy is to be searched in a space of size 2. To illustrate the concept, we take a matrix-vector product $y_{\Psi \Delta} = A_{\Psi} x_{\Delta}$, in which the entries of $y, A$ and $x$ depend on $\Psi, \Delta \subset \Lambda$. Let us assume that $y_{\Psi \Delta}$ itself is an operand of a larger scalar-valued expression, so its entries can represent any possible sub-expression in $\lbrace \alpha_{eij}, \beta_{eik}, \sigma_{ei} \rbrace$. For simplicity, assume that $A$ is the inverse Jacobian matrix of a given coefficient and $x$ the gradient of a function in two dimensions, and consider the case $\Psi = \emptyset$, $\Delta = \lbrace L_i \rbrace$. Thus an entry in $y_{L_i}$ is of the form $(a f_i + b g_i) (c f_i + d g_i) \in \sigma_{ei}$. To schedule this expression, we basically have two options:
\begin{enumerate}
\item We leave the expression unmodified in the body of $L_i$ (in general, one may consider applying generalized code motion or creating a temporary for CCSE -- but it is not the case of this simple example).
\item We notice the presence of spatial sharing. We can recast the expression as $f_i (a + c) + g_i (b + d)$, which promotes temporal sharing and, therefore, code motion. 
\end{enumerate}
The reader can verify that option 2) improves the operation count over 1) if and only if the size of $L_i$ is strictly greater than $c=1$. One can appreciate this dichotomy in a wide range of variational forms\footnote{In fact, it systematically appears in all of the forms used for experimentation (Section~\ref{sec:perf-results}), and many more.}, obviously with varying values of $c$.

In general, given a product of two sub-expressions in a finite element integration loop nest, we observe that whenever there is no ambiguity in the elimination of spatial sharing, the problem of finding an optimal scheduling reduces to the comparison of two alternatives. We note that this condition simply holds if a loop is guaranteed to be linear; in the next section, we will show how to take advantage of this property to ensure quasi-optimal sharing elimination. Further, if the $\sigma_{ei}$ expressions are sufficiently regular, the same approach can be applied to all of the $L_i$-dependent expressions. Structured tensor operations within $\sigma_{ei}$ are not as uncommon as one may think in the first place: for instance, they systematically arise in the weak variational form of the complex hyperelastic model analyzed in Section~\ref{sec:perf-results} (e.g., the Cauchy-Green tensor).

@@

\subsubsection{Algorithm}
\label{sec:sec:se-algo}
Algorithm~\ref{algo:opt} details the steps for performing sharing elimination on a tree representation of the loop nest. The algorithm is meant to be applied to each statement appearing in the independent basic block of the input finite element integration loop nest. It exploits the observation that test and trial functions are always part of structured tensor operations, due to the nature of finite element integration expressions.

\begin{Algo}[Sharing elimination]
\label{algo:opt}
\normalfont
The initialization step consists of a depth-first visit of the expression tree with recursive collection of identical sub-expressions, to ensure the uniqueness of operands. For example, the expression $(a + a + b + c + a)$ is transformed into $(3a + b + c)$.

We perform a second depth-first visit of the expression tree to collect information about multilinear ``symbols'' and ``operands''. For example, any sub-expression $\alpha_{eij}$ ($\beta_{eik}$) is regarded as an operand, while a symbol could be any test (trial) function (or derivatives of) $a_{ij} \in \alpha_{eij}$. In particular, for each symbol $s$, we store a 2-tuple ${<}n,\ \lbrace ops \rbrace{>}$, where $n$ is the number of occurrences of $s$ in the input expression, while $\lbrace ops \rbrace$ represents the set of operands in which $s$ appears.

Let $\mathbb{O}$ be the set of all $\lbrace ops \rbrace$s.

%ops\ =\ \lbrace \alpha_{eij}^{1}, ..., \alpha_{eij}^{N}\rbrace$

%and we recursively label each node $n$ with a 2-tuple ${<}I_{jk},\ op_{gcm}{>}$, where: $I_{jk}$ represents the intersection of the operands depending on the multilinear loop nest within $n$, while $op_{gcm}$ denotes the gain in operation count if generalized code motion were applied to all sub-expressions rooted in $n$ (i.e., if they were placed sufficiently far in the loop nest such that redundant computation were avoided). A crucial observation is that $I_{jk}$ is, in practice, always very small (a few units), even if the form is complex and includes several monomials. In addition, its computation can be made extremely efficient by also tracking the union of the symbols rooted in $n$.
%
%On a subsequent depth-first visit, we examine the label of each operator \texttt{+}. If $I_{jk} = \emptyset$ -- that is, no spatial sharing -- we just proceed to the parent node. Otherwise, we speculatively assume that structured tensor operations are being performed in the sub-tree rooted in \texttt{+}, and compare the operational cost of the two alternatives described in Section~\ref{sec:se-rln} (see points (1) and (2)). Alternative (1) has an operation count equal to $op_{gcm}$. Alternative (2) requires collecting the shared symbols and estimating the impact of generalized code motion on the new factorization (similarly to Figure~\ref{code:multi_loopnest}). For this, we need decide the collection order; that is, whether to start with symbols in $I_j$ or $I_k$. It can easily be demonstrated by contradiction that starting with $I_o$ such that $|L_o| |I_o|$ is minimum ($o \in {\lbrace} j, k{\rbrace}$, with $|.|$ representing a generic ``\texttt{sizeof}'' operator) leads to the lowest operation count. Once all of the operands rooted in Alternative (2) is immediately applied if determined to be ; alternative (1) is instead applied only after all operands rooted in \texttt{+} have been examined.
